---
title: "How to build tables with gt"
linkTitle: How to build tables with gt
date: '2022-08-01'
output:
  html_document: default
  pdf_document: default
highlight: yes
type: book
---

In publications, tables are extremely important to compare individual values, while plots are for showing traces or patterns. The *gt* package, can generate quality tables in R. Here I am going to show you a single line of code in order to produce a table. In addition the complexity of our code will be increased by adding a special design to our table, such as changes in fonts, in colors and footnotes. We will go ahead and we will use the the same data-set from the *Basic Data Wrangling* blog content.

## Downloading Data

```{r, echo=FALSE, results='hide'}
library(downloadthis)
```

```{r, echo=FALSE}
download_link(
  link = "https://github.com/andvenizelos/MyBlog_R_For_Biologists/blob/main/Tip.1/Encode_HMM_data.txt",
  button_label = "Download txt file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

## Install Packages and Loading Libraries

First step is to install packages either from CRAN, Bioconductor or from Github and to load the libraries.

```{r, warning=FALSE, message=FALSE}
library(data.table)
library(dplyr)
library(tidyverse)
library(plyr)
library(scales)
library(tidyquant)
library(gt)
```

## Importing Data

Navigate your directory to your folder where you have saved the data set.

```{r, echo=TRUE}

filename <- "~/Documents/MyWebPage/content/blog/example/Encode_HMM_data.txt"

```

Now we will go ahead to import our data set using the **fread()** function.

```{r, echo = TRUE}
my_data <- fread(filename)
```

Then you need to polish your dataset and split into a dataframe by chromosome and type with **ddply()** function from *plyr* package.

```{r, echo=TRUE}
colnames(my_data)[1:4] <- c("chrom","start","stop","type")
my_data <- my_data[,c(1:4)] #Keeping only the columns 1 to 4
my_data$type <- gsub('[[:digit:]]+_','',my_data$type) #removing the numbers and underscores from our type column
my_data2 <- ddply(my_data, .(type, chrom), summarise , no = length(type))
glimpse(my_data2)
```

## Pivot_wider

**pivot_wider()** makes the data-set **wider** and is often utilized to tidy long data sets and required for the purpose of creating tables.

```{r, echo=TRUE}
my_data_final <- my_data2 %>% group_by(type) %>% 
  mutate(prop = no/sum(no)) %>% 
  ungroup() %>% 
  pivot_wider(
    id_cols = type,
    names_from = chrom,
    values_from = prop
  )

```

```{r, echo=TRUE}
head(my_data_final)
```

## Data Wrangling

Our data frame is almost ready to be converted into a table. However, we need a bit further to polish our data in order to add a heatmap to our cells illustrating differences in the range of specific values for Chr19. Below I am showing a simple way of doing this using the function **order**.

```{r, echo=TRUE, warning=FALSE}

my_data_final <- my_data_final[order(my_data_final[,12],decreasing = TRUE),]

```

## Create a table

Finally we are ready to convert our dataset into a table by running a single line of code.

```{r, echo=TRUE}
table_gt <- my_data_final %>% gt()
table_gt
```

## Table Options

Our table is generated, however it needs further customization. Few finer details such as tab style, font and colors can be changed using the **opt_xxx()** and **tab_options()** functions. In addition you can also tweak values in the data_frame using the **fmt_number()** function. In our dataset due high amount of decimals we restricted down to five decimals for each of the values.

```{r, echo=TRUE}
table_gt_final <- table_gt %>%  opt_all_caps() %>% 
  opt_table_font(
    font = list(
      google_font("Lato"),
      default_fonts()
    )
  ) %>% 
  opt_row_striping() %>% 
  tab_options(
    row.striping.background_color = "#fafafa",
    table_body.hlines.color = "#f6f7f7",
    source_notes.font.size = 12,
    table.font.size = 16,
    table.width = px(700),
    heading.align = "left",
    heading.title.font.size = 24,
    table.border.top.color = "transparent",
    table.border.top.width = px(3),
    data_row.padding = px(7)
  ) %>% fmt_number(
    columns = 2:24,
    decimals = 5,
  use_seps = TRUE
  ) %>% 
  tab_style(style = list(
    cell_text(style = "italic")),
    locations = cells_body(
      columns = c(type)
    ))
```

```{r}
table_gt_final
```

## Adding some style and colour

Our table has already got a better design, however as we mentioned above we will add a heatmap to our cells with a color showing the differences in numneric values (min and max) for chr19. To do that we need to create a palette and apply conditional colouring using the function **data_color()**.

```{r}
min_Chr19 <- min(my_data_final$chr19)
max_Chr19 <- max(my_data_final$chr19)
palette <- col_numeric(c("#FEF0D9", "#990000"), domain = c(min_Chr19, max_Chr19), alpha = 0.75)
               
table_gt_final %>% data_color(columns = c(chr19),
               colors = palette) 
```

## Adding Footnotes

Footnotes are important for explanatory (content) notes or copyright permission. Therefore footnotes could also be significant for academic/scientific tables with more complex data-sets. Footnotes are usually placed at the bottom of the table. One way to add a footnote with *gt* package is to use the function **tab_footnote()**. In addition you have the option to modify the set of footnote marks with the fucntion **opt_footnote_marks**. The valid keywords are : **numbers** (for numeric marks), **letters** (for alphabetic marks), **standard** (for traditional symbol marks) and **extended** (which adds more symbols to the standard set.) For our table we have used the **numbers** keyword as an option for footnote mark.

```{r}
table_gt_final2 <- table_gt_final %>% data_color(columns = c(chr19),
               colors = palette) %>% 
               tab_footnote(
               footnote = md("Enhancers"),
               locations = cells_body(
               columns = c("type"),
               rows = c(1,3))) %>% 
  tab_footnote(footnote = md("Promoters"),
               locations = cells_body(
                 columns = c("type"),
                 rows = c(4,8, 12)
                 )) %>% opt_footnote_marks(marks = "numbers")
table_gt_final2
```

## Save the table

And now we are ready to save our table. The **gtsave()** function makes it easy to save a **gt** table to a file in a different format such as (HTML, PDF, PNG, LaTeX, or RTF file) via the {webshot} package.

```{r,warning=FALSE, cache.comments=FALSE,results='hide'}
webshot::install_phantomjs() 
table <- table_gt_final2 %>% gtsave("~/Documents/MyWebPage/content/blog/example/table.pdf")
```

## Feel free to download the files from github: 

<a id="raw-url" href="https://github.com/andvenizelos/MyBlog_R_For_Biologists/tree/main/Tip.5">Download FILES</a>