---
date: "2022-05-20"
linkTitle: How to create a Heatmap
title: How to create a Heatmap
type: book
highlight: TRUE
output: html_document
weight: 30
---

```{r,warning=FALSE, message=FALSE, echo=FALSE}
library(downloadthis)
```

## Downloading Data

For this session we are going to utilize the dataset from the <https://encode.project.org/>.

You can download the **txt file** here:

```{r, echo=FALSE}
download_link(
  link = "https://github.com/andvenizelos/MyBlog_R_For_Biologists/blob/main/Tip.1/Encode_HMM_data.txt",
  button_label = "Download txt file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

## Install Packages and Loading Libraries

First step is to install packages either from CRAN, Bioconductor or from Github and to load the libraries.

```{r, warning=FALSE, message=FALSE}
library(data.table)
library(dplyr)
library(tidyverse)
library(plyr)
library(scales)
library(tidyquant)
```

## Importing Data
Navigate  your directory to your folder where you have saved the data set.
Now we will go ahead to import our data set using the *fread* function.

```{r, echo=FALSE, results='hide'}
setwd("~/Documents/Running_Projects/MyWebPage/content/blog/example/")
filename <- "~/Documents/Running_Projects/MyWebPage/content/blog/example/Encode_HMM_data.txt"

```

```{r}
my_data <- fread(filename)
```

## Data wrangling
In order to have a clean data set we need to add column names and to remove the columns that are not needed for our final heatmap output.

```{r}
colnames(my_data)[1:4] <- c("chrom","start","stop","type")
my_data <- my_data[,c(1:4)] #Keeping only the columns 1 to 4
my_data$type <- gsub('[[:digit:]]+_','',my_data$type) #removing the numbers and underscores from our type column
glimpse(my_data)
```

## Split the data frame
We can now split our data frame by chromosome and type and apply a function summarizing the length for each type per chromosome using the **ddply** function from the **plyr** library.
```{r}
my_data2 <- ddply(my_data, .(type, chrom), summarise , no = length(type))
glimpse(my_data2)

```

## Pivot_longer & Pivot_wider
__pivot_longer()__ makes dataset **longer** by increasing the number of rows and decreasing the number of columns, whereas __pivot_wider()__ is often utilized to tidy long data sets and often required for the purpose of the analysis (for example creating a heatmap)

```{r}
my_data_final <- my_data2 %>% group_by(type) %>% 
  mutate(prop = no/sum(no)) %>% 
  ungroup() %>% 
  pivot_wider(
    id_cols = type,
    names_from = chrom,
    values_from = prop
  ) %>% arrange(-`chr19`) %>% 
  mutate(type = fct_reorder(type, `chr19`)) %>% 
  pivot_longer(
    cols = -type,
    names_to = "chrom",
    values_to = "prop"
  )

head(my_data_final)
```

## Creating a Heatmap

Finally we are ready to generate our first heatmap! One way to do that is to use the **ggplot** library.

```{r}
heatmap <- my_data_final %>% 
ggplot(aes(chrom, type)) +
  geom_tile(aes(fill = prop)) +
  geom_text(aes(label = scales::percent(prop, accuracy = 1)),
            size = 3) +
  scale_fill_gradient(low = "white",high = palette_light()[1]) +
  labs(
    title = "Percentage of each of the types per chromosome",
    x = "Chromosomes",
    y = "Type"
  )+
  theme_tq() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(face = "bold"))
    
  

    
heatmap


```
