---
date: "2022-06-23"
linkTitle: How to read multiple csv files
title: How to read multiple csv files
type: book
highlight: TRUE
output: html_document
weight: 10
---

```{r,warning=FALSE, message=FALSE, echo=FALSE}
library(downloadthis)
```

## Downloading Data

For this session we are going to utilise the dataset from the <https://encode.project.org/>.

You can download the **txt file** here:

```{r, echo=FALSE}
download_link(
  link = "https://github.com/andvenizelos/MyBlog_R_For_Biologists/blob/main/Tip.1/Encode_HMM_data.txt",
  button_label = "Download txt file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

## Install Packages and Loading Libraries

First step is to install packages either from CRAN, Bioconductor or from Github and to load the libraries.

```{r, warning=FALSE, message=FALSE}
library(data.table)
library(dplyr)
library(tidyverse)
library(plyr)
library(fs)
```

## Importing Data

Navigate your directory to your folder where you have saved the data set. Now we will go ahead to import our data set using the fread command.

```{r, echo=FALSE, results='hide'}
setwd("~/Documents/Running_Projects/MyWebPage/content/blog/example/")
filename <- "~/Documents/Running_Projects/MyWebPage/content/blog/example/Encode_HMM_data.txt"

```

```{r}
my_data <- fread(filename)
```

## Data wrangling

In order to have a clean data set we need to add column names and to remove the columns that are not needed for our final heatmap output.

```{r}
colnames(my_data)[1:4] <- c("chrom","start","stop","type")
my_data <- my_data[,c(1:4)] #Keeping only the columns 1 to 4
my_data$type <- gsub('[[:digit:]]+_','',my_data$type) #removing the numbers and underscores from our type column
glimpse(my_data)
```

## Creating a new directory

We can now create a new directory where we can save our *.csv* files.

```{r}
new_directory <- "~/Documents/Running_Projects/MyWebPage/content/blog/example/my_csv_files/"
dir_create(new_directory)
```

## Spliting our dataframe

We are now ready to split our data frame into multiple *.csv* files and we can save them under the directory **/my_csv_files**.

```{r, echo=TRUE, results='hide',warning=FALSE}
setwd("~/Documents/Running_Projects/MyWebPage/content/blog/example/my_csv_files/")
my_data %>%
    group_by(chrom) %>%
    group_split() %>%
    map(
        .f = function(data) {
            write_csv(data, path = unique(data$chrom))
        }
    )
```

## Importing multiple csv files [1]

We can do this in multiple ways, either using the **lapply** function after generatiing a list or the **map** function. Let's go ahead and use the first way :

```{r}
setwd("~/Documents/Running_Projects/MyWebPage/content/blog/example/my_csv_files/")
fileslist = list.files(pattern = "")
csvFiles = lapply(fileslist, function(x)read.table(x, header = T, sep = ","))
csvFiles = do.call("rbind", csvFiles)
csvFiles |> head()
```

## Importing multiple csv files [2]

And the second way is to import the multiple csv files from our directory using the *.map* function.

```{r}
directory_that_holds_files <-("~/Documents/Running_Projects/MyWebPage/content/blog/example/my_csv_files/")

chromosomes_list <- directory_that_holds_files %>%
  dir_ls() %>%
  map(
    .f = function(path)read.table(path, header = T, sep = ","))

```

## Binding Rows

And the final step is to bind the data frame with the *bind_rows* function.

```{r}
chromosomes_tbl <- chromosomes_list %>%
    set_names(dir_ls(directory_that_holds_files)) %>%
    bind_rows(.id = "file_path")
head(chromosomes_tbl)

```
### You can find the csv split files here : 


<a id="raw-url" href="https://github.com/andvenizelos/MyBlog_R_For_Biologists/tree/main/Tip.4">Download FILES</a>


