---
date: "2022-03-30"
linkTitle: Data Wrangling
title: Basic Data Wrangling
type: page
output: html
---

```{r,warning=FALSE, message=FALSE, echo=FALSE}
library(downloadthis)
```
## Downloading Data
One of the most important lesson in R is filtering and cleaning up you data, in order to generate an easy to read and to handle dataset. First of all we need to import a dataset. There are many sources for publicly avaliable big data such as: <https://genome.ucsc.edu/> , <https://www.ensembl.org/>, <https://www.encodeproject.org/>. I have already downloaded a dataset for you. You can download the **txt file** here:

```{r, echo=FALSE}
download_link(
  link = "https://github.com/andvenizelos/MyWebPage/tree/main/Lessons/Encode_HMM_data.txt",
  button_label = "Download txt file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```
## Install Packages and Loading Libraries

First step is to install packages either from CRAN, Bioconductor or from Github and to load the libraries.

```{r, warning=FALSE, message=FALSE}
library(data.table)
library(dplyr)
library(tidyverse)
library(tm)
```

## Importing Data

Now we will go ahead to import our dataset. There are many ways to import a dataset depending on the file type and the size of the file.
First step is to navigate your console to your own directory with your folder containing the data below.
Here I am going to show you two ways of importing a dataset with the latter one to be preferred for larger datasets.

### Method 1

```{r}
filename <- "Encode_HMM_data.txt"
my_data <- read.csv(filename, sep="\t", header=FALSE)
```

### Method 2

```{r}
library(data.table)
my_data <- fread(filename)
```

```{r}
head(my_data)
```

```{r}
glimpse(my_data)
```


## Select Columns

We can see that succesfully we have imported our dataset in R studio. It's a dataframe with *571,339* rows and *9* columns. We will go ahead and we will select the first *4* columns from our data set and to ignore the rest of the columns. Again there are many ways of doing that. Here I can show you *2* ways :

### Method 1

```{r}
my_data %>% select(V1,V2,V3,V4)
```

### Method 2

``` {r}
my_data[,1:4]
```


``` {r}
sorted_data <- my_data[,1:4]
```
## Adding Column Names

Now we have selected successfully the columns that we need. However, we need to rename the columns in order to have a more clean data set.

``` {r}
colnames(sorted_data)[1:4] <- c("Chromosome","Start","Stop","Type")
head(sorted_data)
```


## Final Clean-Up

Now we have a more defined clean data set. However, there is a string including a number and an underscore as prefix in each character value of the column *Type*. Below again *2* ways to remove this string:

### Method 1

```{r}
sorted_data$Type <- gsub('[[:digit:]]+_','', sorted_data$Type)
```

### Method 2

``` {r}
sorted_data$Type <- str_replace_all(sorted_data$Type, "[:digit:]+_","")
head(sorted_data)
```


And here you have it ! A final clean data set easy to read and work with.
